### 2025-08-18 chapters/benchmark.tex
conclusion finished

    % Second Subfigure: Enclave Run10_SEQ vs Run5*
        \begin{subfigure}{\textwidth}
        \centering
        \begin{tikzpicture}
        \begin{axis}[
            ybar,
            bar width=8pt,
            width=\textwidth,
            height=8cm,
            ymode=log,
            log origin=infty,
            ymin=0.001,
            ylabel={Value (log scale)},
            symbolic x coords={
                Total\ Time,
                Iterations/Sec,
                User\ CPU\ Time,
                System\ CPU\ Time,
                Max\ RSS,
                Minor\ PF,
                Major\ PF,
                Voluntary\ CS,
                Involuntary\ CS
            },
            xtick=data,
            x tick label style={rotate=30, anchor=east},
            enlarge x limits=0.05,
            ymajorgrids,
            grid style=dashed,
            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
            nodes near coords,
            every node near coord/.append style={
                font=\tiny,
                rotate=90,
                anchor=west,
                /pgf/number format/.cd, fixed, precision=2
            }
        ]
        % Native (Run10_SEQ)
        \addplot[pattern=north east lines,pattern color=black] coordinates {
            (Total\ Time,101.11)
            (Iterations/Sec,989.00)
            (User\ CPU\ Time,101.14)
            (System\ CPU\ Time,0.01)
            (Max\ RSS,1691.60)
            (Minor\ PF,83.60)
            (Major\ PF,0.20)
            (Voluntary\ CS,0.90)
            (Involuntary\ CS,350.40)
        };
        % Enclave (Run10_SEQ)
        \addplot[pattern=dots,pattern color=black] coordinates {
            (Total\ Time,107.02)
            (Iterations/Sec,929.80)
            (User\ CPU\ Time,41.16)
            (System\ CPU\ Time,41.00)
            (Max\ RSS,10.00)
            (Minor\ PF,1.00)
            (Major\ PF,0.001) % zero replaced by small non-zero
            (Voluntary\ CS,49.00)
            (Involuntary\ CS,65992.00)
        };
        \legend{Native (Run10\_SEQ), Enclave (Run10\_SEQ)}
        \end{axis}
        \end{tikzpicture}
        \caption{Log-scale comparison of key metrics for CoreMark 64M\_1Core: Native vs Enclave (Run10\_SEQ).}
        \label{fig:coremark_native_vs_enclave_log}
    \end{subfigure}

>>>>

    % Second Subfigure: Enclave Run10_SEQ vs Run5*
\begin{subfigure}{\textwidth}
    \centering
    \begin{tikzpicture}
    \begin{axis}[
        ybar,
        bar width=8pt,
        width=\textwidth,
        height=8cm,
        ymode=log,
        log origin=infty,
        ymin=0.001,
        ylabel={Value (log scale)},
        symbolic x coords={
            Total\ Time,
            Iterations/Sec,
            User\ CPU\ Time,
            System\ CPU\ Time,
            Max\ RSS,
            Minor\ PF,
            Major\ PF,
            Voluntary\ CS,
            Involuntary\ CS
        },
        xtick=data,
        x tick label style={rotate=30, anchor=east},
        enlarge x limits=0.05,
        ymajorgrids,
        grid style=dashed,
        legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
        nodes near coords,
        every node near coord/.append style={
            font=\tiny,
            rotate=90,
            anchor=west,
            /pgf/number format/.cd, fixed, precision=2
        }
    ]
    % Native (Run10_SEQ average)
    \addplot[pattern=north east lines,pattern color=black] coordinates {
        (Total\ Time,100.91)
        (Iterations/Sec,989.80)
        (User\ CPU\ Time,100.99)
        (System\ CPU\ Time,0.013)
        (Max\ RSS,1691.60)
        (Minor\ PF,83.70)
        (Major\ PF,0.20)
        (Voluntary\ CS,0.90)
        (Involuntary\ CS,350.40)
    };
    % Enclave (Run10_SEQ average)
    \addplot[pattern=dots,pattern color=black] coordinates {
        (Total\ Time,107.45)
        (Iterations/Sec,928.97)
        (User\ CPU\ Time,41.16)
        (System\ CPU\ Time,41.00)
        (Max\ RSS,10.00)
        (Minor\ PF,1.00)
        (Major\ PF,0.001) % zero replaced by small non-zero for log scale
        (Voluntary\ CS,49.00)
        (Involuntary\ CS,65992.00)
    };
    \legend{Native (Run10\_SEQ avg), Enclave (Run10\_SEQ avg)}
    \end{axis}
    \end{tikzpicture}
    \caption{Log-scale comparison of key metrics for CoreMark 64M\_1Core: Native vs Enclave (averaged over 10 runs).}
    \label{fig:coremark_native_vs_enclave_log}
\end{subfigure}

### 2025-08-18 chapters/conclusion.tex
updated TOC
conclusion finished

\section{Summary of Contributions}

\section{Limitations of the Current Study}

\section{Directions for Future Research}

>>>>

\chapter{Conclusion}
\label{chap:conclusion}

The research presented in this thesis has provided a comprehensive examination of the performance, scalability, and architectural constraints involved in deploying secure enclaves on RISC-V platforms through the Keystone Trusted Execution Environment framework. By conducting rigorous performance benchmarking, detailed microarchitectural analyses, and thorough system-level evaluations, this study has elucidated both the enabling strengths and the critical limitations inherent in current RISC-V TEE implementations, with particular focus on the Physical Memory Protection mechanism responsible for enclave isolation.

Through the empirical evaluation of a range of workloads—including compute-bound benchmarks and memory-intensive cryptographic algorithms—this thesis has demonstrated that enclave performance on RISC-V is highly workload-dependent. Compute-bound applications such as CoreMark and Dhrystone experience relatively minor performance degradation when executed within enclaves, typically ranging between 10 to 15 percent. This observation aligns with expectations given the dominance of predictable integer computations and limited memory operations in such workloads. In stark contrast, cryptographic workloads exemplified by the Kyber post-quantum algorithm suffered from substantial performance penalties exceeding 60 percent. This discrepancy arises from Kyber's memory-bound nature and its frequent indirect memory accesses, which exacerbate the overhead introduced by enclave memory isolation and associated integrity checks.

Beyond performance considerations, a critical bottleneck was identified in the utilization of PMP for enclave memory protection. The RISC-V architecture’s limit of sixteen PMP entries, each constrained by the Naturally Aligned Power-of-Two encoding scheme, imposes stringent limits on enclave scalability. Fragmented or non-contiguous memory allocations rapidly exhaust PMP entries, restricting the feasible number of concurrent enclaves. Empirical data confirmed that enclave instantiation failures occurred beyond eight simultaneous enclaves due to PMP saturation, a constraint that holds even on systems with abundant CPU cores and physical memory. This architectural limitation presents a significant barrier to multi-tenant or high-concurrency scenarios where isolated execution contexts must coexist.

The implications of these findings extend beyond immediate performance and capacity concerns. The current PMP-based isolation model, while offering lightweight hardware-enforced security guarantees, intrinsically restricts flexibility in enclave memory layout and scalability. Without modifications to either the hardware architecture—such as increasing PMP register counts or adopting more flexible memory region descriptors—or to system software managing memory allocation, the practicality of RISC-V enclaves in complex real-world applications remains limited.

To mitigate these challenges, this thesis advocates several best practices. Firstly, contiguous and properly aligned memory allocation is essential to minimize PMP entry consumption and reduce fragmentation. Secondly, consolidating multiple lightweight tasks into fewer enclaves can relieve PMP pressure. Thirdly, selectively isolating only the most security-critical components within enclaves, particularly in cryptographic workloads, balances security with performance. Furthermore, runtime systems could implement dynamic PMP entry management, reclaiming and reconfiguring PMP resources in response to enclave lifecycle events, though this approach demands additional synchronization mechanisms to preserve security guarantees. Lastly, optimizing enclave entry and exit frequency, particularly within tightly looped cryptographic operations, can help amortize synchronization costs.

Looking forward, this research underscores several promising directions. Architecturally, enhancements to RISC-V PMP—such as increasing the number of PMP entries or introducing hierarchical and flexible region descriptors—could substantially improve enclave scalability. From a software perspective, enclave-aware memory allocators and scheduling systems that proactively manage PMP resources and workload characteristics would enhance system robustness under high concurrency. Integration of cryptographic accelerators within the TEE boundary also holds promise to alleviate performance bottlenecks inherent in memory-bound cryptographic workloads. Compiler toolchains and enclave abstraction layers that enforce memory alignment and contiguity at build time could reduce developer burden and improve runtime efficiency.

In summary, this thesis has demonstrated that while RISC-V enclaves leveraging Keystone represent a compelling foundation for open and customizable trusted execution environments, practical deployment necessitates careful consideration of PMP-imposed constraints. Balancing security, performance, and scalability demands architectural awareness and judicious system design. The insights gained herein provide a detailed characterization of enclave behavior and highlight fundamental limitations, serving both as a guide for current practitioners and as a foundation for future research in secure computing on open hardware platforms. As trusted execution environments become increasingly vital across cloud, edge, and IoT domains, addressing these challenges will be critical to realizing scalable, efficient, and secure computing paradigms.

%6.1 Summary of Findings
%6.2 Limitations
%6.3 Directions for Future Research

### 2025-08-18 chapters/discussion.tex
conclusion finished

\section{Implications for Secure System Design and Cryptographic Workloads}
\label{sec:implications}
The challenges highlighted in the previous section, particularly those related to the RISC-V Physical Memory Protection (PMP) mechanism, have significant implications for the design of secure systems, especially those that need to handle cryptographic workloads. As the demand for secure enclaves continues to grow—particularly in cloud environments, edge computing, and multi-tenant systems—the trade-offs between security and performance must be carefully balanced. While the ability to enforce strong memory isolation using PMP is critical for ensuring the confidentiality and integrity of sensitive data, it also introduces system bottlenecks that can severely impact performance, particularly for memory-bound tasks such as cryptography.

@@

Furthermore, the trade-off between memory usage and security introduces additional complexity in system design. Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme. For cryptographic workloads that often require non-contiguous memory layouts, the inability to efficiently manage PMP entries could lead to further fragmentation and resource wastage. This increases the overall system overhead and reduces the number of concurrent enclaves that can be supported.

For secure systems to scale effectively in the face of these limitations, both hardware and software-level strategies must be employed. On the hardware side, increasing the number of available PMP entries or providing more flexible encoding schemes would alleviate some of the bottlenecks. However, these hardware solutions would require significant architectural changes and may not be immediately feasible in current RISC-V processors.

On the software side, dynamic management of PMP entries, enclave multiplexing, and memory layout optimization represent viable paths forward. By reusing PMP entries or consolidating multiple lightweight workloads into a single enclave, the overall resource consumption can be reduced. Additionally, optimizing the memory layout to fit within the NAPOT constraints can help minimize fragmentation, thereby reducing the number of PMP entries required per enclave. These software-level strategies can help mitigate some of the scalability issues but may come at the cost of added complexity in the system design.

\begin{figure}[htbp]
\centering
% First subfigure (larger)
\begin{minipage}{0.9\textwidth}
\centering
\begin{tikzpicture}[scale=1, transform shape]
\begin{axis}[
    ybar,
    bar width=10pt,
    width=\linewidth, height=6cm,
    ylabel={PMP entries required},
    symbolic x coords={128\,KB aligned,192\,KB misaligned,300\,KB misaligned,384\,KB misaligned},
    xtick=data,
    x tick label style={rotate=20, anchor=east},
    ymajorgrids, grid style=dashed,
    enlarge x limits=0.15,
    legend style={at={(0.5,-0.3)},anchor=north,legend columns=2},
    pattern color=black,
    nodes near coords, every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
]
\addplot[pattern=north east lines] coordinates {
(128\,KB aligned,1)
(192\,KB misaligned,2)
(300\,KB misaligned,2)
(384\,KB misaligned,1)
};
\addplot[pattern=crosshatch] coordinates {
(128\,KB aligned,2)
(192\,KB misaligned,3)
(300\,KB misaligned,3)
(384\,KB misaligned,2)
};
\legend{Contiguous mapping,Fragmented mapping}
\end{axis}
\end{tikzpicture}
\caption*{(a) Fragmentation and misalignment increase PMP entry usage}
\label{fig:pmp-fragmentation-entries}
\end{minipage}

\vspace{1.5em}

% Second subfigure (also larger)
\begin{minipage}{0.9\textwidth}
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]
% Memory bar
\draw[thick] (0,0) rectangle (12,1);
\node[anchor=south] at (0,1.1) {Base};
\node[anchor=south] at (12,1.1) {Top};

@@

\node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
\end{tikzpicture}
\caption*{(b) Fragmentation overhead: non-contiguous regions mapped to NAPOT-aligned blocks, with hatched padding indicating waste}
\label{fig:napot-fragmentation}
\end{minipage}

\caption{Impact of fragmentation and alignment on PMP entry utilization and memory layout.}
\label{fig:pmp-fragmentation-group}
\end{figure}

@@

\section{Configuration Best Practices and Deployment Recommendations for RISC-V TEEs}

Deploying Trusted Execution Environments (TEEs) on RISC-V platforms using the Keystone framework presents a unique set of challenges and opportunities that must be carefully navigated to achieve optimal security and performance. Based on the extensive performance benchmarking and system analysis conducted in Chapters~\ref{chap:benchmarking} and \ref{chap:discussion}, several configuration best practices emerge, offering practical guidance for system architects, developers, and security engineers seeking to deploy enclaves across a variety of workload types and operational contexts.

@@

>>>>

\section{Implications for Secure System Design and Cryptographic Workloads}
\label{sec:implications}

The challenges highlighted in the previous section, particularly those related to the RISC-V Physical Memory Protection (PMP) mechanism, have significant implications for the design of secure systems, especially those that need to handle cryptographic workloads. As the demand for secure enclaves continues to grow—particularly in cloud environments, edge computing, and multi-tenant systems—the trade-offs between security and performance must be carefully balanced. While the ability to enforce strong memory isolation using PMP is critical for ensuring the confidentiality and integrity of sensitive data, it also introduces system bottlenecks that can severely impact performance, particularly for memory-bound tasks such as cryptography.

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape]
\begin{axis}[
    ybar,
    bar width=10pt,
    width=0.9\linewidth, height=6cm,
    ylabel={PMP entries required},
    symbolic x coords={128\,KB aligned,192\,KB misaligned,300\,KB misaligned,384\,KB misaligned},
    xtick=data,
    x tick label style={rotate=20, anchor=east},
    ymajorgrids, grid style=dashed,
    enlarge x limits=0.15,
    legend style={at={(0.5,-0.3)},anchor=north,legend columns=2},
    pattern color=black,
    nodes near coords, every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
]
\addplot[pattern=north east lines] coordinates {
(128\,KB aligned,1)
(192\,KB misaligned,2)
(300\,KB misaligned,2)
(384\,KB misaligned,1)
};
\addplot[pattern=crosshatch] coordinates {
(128\,KB aligned,2)
(192\,KB misaligned,3)
(300\,KB misaligned,3)
(384\,KB misaligned,2)
};
\legend{Contiguous mapping,Fragmented mapping}
\end{axis}
\end{tikzpicture}
\caption{Fragmentation and misalignment increase PMP entry usage, showing how non-ideal alignment leads to more PMP entries per enclave.}
\label{fig:pmp-fragmentation-entries}
\end{figure}

To address overhead and scalability challenges, **enclave multiplexing** offers a promising software-level strategy by consolidating multiple lightweight workloads into a single enclave. This reduces the number of PMP entries required by reusing them across tasks, optimizing resource usage and improving scalability. An example of a multiplexed enclave task dispatcher is shown in Listing~\ref{lst:enclave-mux}.

\begin{lstlisting}[language=Python, caption={Multiplexed enclave task dispatcher.}, label={lst:enclave-mux}]
def enclave_dispatcher(task_id, input_data):
    if task_id == 1:
        return process_crypto(input_data)
    elif task_id == 2:
        return handle_key_store(input_data)
    elif task_id == 3:
        return run_diagnostics(input_data)
    else:
        raise ValueError("Unknown task")
\end{lstlisting}

Furthermore, the trade-off between memory usage and security introduces additional complexity in system design. Enclaves need to be carefully sized and aligned to fit within the constraints of PMP's encoding scheme, particularly the NAPOT (Naturally Aligned Power-Of-Two) encoding used to specify memory regions. For cryptographic workloads that often require non-contiguous memory layouts, inefficient PMP entry management leads to fragmentation and resource wastage, increasing overhead and limiting concurrent enclaves.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=1, transform shape,font=\small]
% Memory bar
\draw[thick] (0,0) rectangle (12,1);
\node[anchor=south] at (0,1.1) {Base};
\node[anchor=south] at (12,1.1) {Top};

@@

\node[anchor=west] at (0,-1.5) {\footnotesize Hatched = padding/waste due to power-of-two alignment};
\end{tikzpicture}
\caption{Fragmentation overhead caused by mapping non-contiguous memory regions to NAPOT-aligned blocks. The hatched areas represent wasted padding required for alignment.}
\label{fig:napot-fragmentation}
\end{figure}

The memory layout complexity is also reflected in code structures like the custom memory layout shown in Listing~\ref{lst:memory-layout}, which attempts to align heap, stack, and code regions to NAPOT boundaries.

\begin{lstlisting}[language=C, caption={Custom memory layout aligned to NAPOT for enclave use.}, label={lst:memory-layout}]
struct enclave_memory_layout {
    void* heap_start;     // Aligned to 0x1000
    void* heap_end;
    void* stack_start;    // Aligned to 0x1000
    void* stack_end;
    void* code_region;    // Executable, aligned
};
\end{lstlisting}

At the hardware configuration level, PMP entries must be programmed carefully to respect alignment and permission constraints. Listing~\ref{lst:pmp-napot} illustrates a PMP configuration for a 4KB NAPOT-aligned enclave memory region, enabling read, write, and execute permissions.

\begin{lstlisting}[language=C, caption={Example PMP configuration for a NAPOT-aligned memory region.}, label={lst:pmp-napot}]
#define PMP_R 0x01
#define PMP_W 0x02
#define PMP_X 0x04
#define PMP_NAPOT 0x18

// Configure PMP entry 0 for a 4KB aligned enclave memory region
pmpaddr0 = (0x80000 >> 2);  // Shifted for NAPOT format
pmpcfg0  = PMP_R | PMP_W | PMP_X | PMP_NAPOT;
\end{lstlisting}

@@

\textbf{Note:} The above code snippets and figures are intended as conceptual illustrations to aid understanding. They represent typical scenarios and strategies but are not directly operational without further system integration and adaptation.

\section{Configuration Best Practices and Deployment Recommendations for RISC-V TEEs}
\label{sec:best-practices}
Deploying Trusted Execution Environments (TEEs) on RISC-V platforms using the Keystone framework presents a unique set of challenges and opportunities that must be carefully navigated to achieve optimal security and performance. Based on the extensive performance benchmarking and system analysis conducted in Chapters~\ref{chap:benchmarking} and \ref{chap:discussion}, several configuration best practices emerge, offering practical guidance for system architects, developers, and security engineers seeking to deploy enclaves across a variety of workload types and operational contexts.

@@

\begin{lstlisting}[caption={Best Practices for RISC-V TEE Deployment}, 
                   label={lst:enclave-best-practices}, 
                   basicstyle=\ttfamily\small, 
                   numbers=left, 
                   numberstyle=\tiny, 
                   xleftmargin=2em]
Allocate enclave memory contiguously, NAPOT-aligned, to minimize PMP usage.
Consolidate workloads into fewer enclaves to reduce resource contention.
Limit active enclaves to prevent PMP exhaustion and creation failures.
Offload non-sensitive tasks to the untrusted domain to lower overhead.
Batch enclave calls to amortize context-switch costs.
Use dynamic PMP management to optimize resources at runtime.
\end{lstlisting}

### 2025-08-18 main.tex
added baseline results
introduced latex diagrams
4.4 complete
5.3 complete
native vs enclave latex plots added
conclusion finished

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{pagella contrast design}
\usepackage{comment}
\usepackage{float}
\lstset{basicstyle=\ttfamily}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, shadows.blur}

>>>>

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{pagella contrast design}
\usepackage{comment}
\usepackage{fontspec}
\usepackage{float}
\lstset{basicstyle=\ttfamily}
\usepackage{tikz}
\usetikzlibrary{matrix}
\usepackage{pgfplots}
\usetikzlibrary{pgfplots.fillbetween}
\pgfplotsset{compat=1.18}
\usepackage{pgfplotstable}
\usepackage{pgf-umlsd}
\usepgfplotslibrary{statistics}
\usetikzlibrary{patterns} 
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[flushleft]{threeparttable}
\usepackage{enumitem}
\sloppy
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, shadows.blur}
\usetikzlibrary{positioning,backgrounds,calc,shapes, shapes,arrows.meta}
\usepackage{xcolor}
\lstset{
    basicstyle=\ttfamily\small,
    frame=single,
    columns=fullflexible,
    breaklines=true,
    backgroundcolor=\color{gray!5},
    keywordstyle=\color{blue!70!black},
    commentstyle=\color{gray!60},
    stringstyle=\color{red!70!black},
    captionpos=b,
    showstringspaces=false,
    tabsize=4
}
